Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-19 manu-draft.Rmd
removed pca analysis

data(rswqdat)
data(rstrnpts)
data(rstrndat)
data(trnsum)
data(allpcadat)

@@

@@

Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparison tests to assess trends between months.  Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release. Further analyses were conducted to compare all data types together using PCA and correlation analyses, to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts. Data describing *K. brevis* cell concentrations were only evaluated qualitatively because the data are from event-based sampling and generally do not represent a random sample appropriate for statistical testing. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.  Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a. For PCA, all variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.  Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.

The "PCA" function from the FactoMineR R package [@Le08] was used for PCA and the "ggord" function from the ggord R package [@Beck21b] was used to plot the results. For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners. Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and evaluated with methods described below. Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.

When appropriate, methods were used from the NADA R package [@Lee20] to accommodate measured concentrations in water quality variables that were below detection.  These included estimates of median, mean, and standard deviation of parameters using the "cenfit" function from NADA.  For the multiple comparison tests, only total nitrogen, secchi, and chl-a data were evaluated, none of which were below detection as noted in the FDEP data.  Nitrate/nitrite concentrations were not evaluated quantitatively because over 80\% of the observations were below detection.  PCA and correlation analyses were based on weekly aggregations by area, using appropriate methods for median estimates with non-detects in the water quality data.

@@

```{r}
expvr <- allpcadat$eig[, 2] %>% 
  round(0)
```

A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates showed significant associations between groups of variables (Figure \@ref(fig:allpca)).  The first principal component axis explained `r expvr[1]`\% of the variation among all variables, whereas the second and third axes explained `r expvr[2]`\% and `r expvr[3]`\% of the remaining variation.  For the water quality variables, the first principal component axis had strong, positive loadings for dissolved oxygen and secchi depth and strong negative loadings for total nitrogen, chl-a, and temperature.  For macroalgae, the first axis had positive loadings with frequency occurrence of red and green macroalgae. For seagrasses, turtle grass (*T. testudinum*) had a strong positive loading with axis 1.  The second principal component axis had strong positive loadings for secchi depth, total phosphorus, and ammonia and negative loadings for salinity and cyanobacteria macroalgae.  The third axis showed a strong negative loading with shoal grass (*H. wrightii*) and positive loading with manatee grass (*S. filiforme*).  Observations by area showed that samples in area 3 were distinct from points in area 1 and grouped by positive loadings on the first PCA axis and negative loadings on the second PCA axis.

Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).  Among the water quality variables, chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH ($p < 0.005$ for all) and negatively correlated with secchi depth ($p < 0.05$). Total nitrogen was positively correlated with total phosphorus, temperature, and ammonia ($p < 0.005$, except $p < 0.05$ for temperature) and negatively correlated with secchi ($p < 0.005$).  Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas red macroalgae was negatively correlated with cyanobacteria ($p < 0.005$).  Among the seagrasses, manatee grass was negatively correlated with shoal grass and turtle grass ($p < 0.005$).  Several notable correlations were also observed between water quality, macroalgae, and seagrass. Chlorophyll-a was negatively correlated with red and green macroalgae ($p < 0.05$), whereas secchi depth was positively correlated with both ($p < 0.05$) and negatively correlated with cyanobacteria ($p < 0.05$).   For water quality and seagrass, chl-a, total nitrogen, and total phosphorus were negatively correlated with turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively correlated with shoal grass ($p < 0.05$).  For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).

@@

Several of the results support known responses of water quality to nutrient loads for shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].  Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations (Figure \@ref(fig:allcor)) and loadings from the PCA results (Figure \@ref(fig:allpca)). Temporally, these changes were observed at different times and for different species of phytoplankton.  The initial increase in chlorophyll-a was first associated with a diatom bloom in April and then by *K. brevis* in June and July.  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where total nitrogen that included organic sources were at high concentrations through April and again peaking in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].

From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability during the summer 2021 in comparison to prior years. Future effects on seagrass persistence in this region of Tampa Bay are unknown at this time, but monitoring continues.  Changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous and not readily apparent over the initial 4 month sampling period.  Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.  Spatially, transects closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay (area 3) had higher frequency occurrence of turtle grass and lower frequency of shoal grass.  This reflects a natural partitioning of the species along a salinity gradient [Figure \@ref(fig:allpca), @Sherwood17].  Correlations between macroalgae and seagrasses can also be explained by spatial differences along a depth gradient. Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.  Shoal grass and turtle grass generally grow at more shallow depths than manatee grass and are often exposed during low tides.  As an adaptive mechanism to varying water levels at shallower depths, blades of shoal grass and turtle grass can "lay down" during low tide to minimize desiccation stress.  Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species grows deeper. In addition to species depth partitioning, these associations may also reflect natural variation in seagrass response to water clarity gradients along the main axis of Tampa Bay.  Shoal grass can occur in more turbid waters of oligohaline areas.

@@

```{r allpca, fig.cap = 'Principal components analysis (PCA) for water quality variables, macroalgae, and seagrass results aggregated by week from April to July 2021 in response to wastewater discharge from Piney Point.  Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed. All variables were standardized to zero mean and unit variance prior to PCA. Variables with log-normal distribution were log-transformed prior to analysis.'}
knitr::include_graphics(here('figs/allpca.jpeg'))
```

>>>>

data(rswqdat)
data(rstrnpts)
data(rstrndat)
data(trnsum)

@@

The historical monitoring data were also used to model an expected seasonal pattern for April to October in 2021.  This was done by estimating smoothed splines using Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay.  GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year. The continuous year value was modeled with a thin plate regression spline and the day of year value was modelled with a cyclic spline [following similar methods @Murphy19].  The modeled results provided an estimate of the expected normal seasonal variation that takes into account a long-term annual trend.  Differences in the observed values sampled in the April to October time periods from the "forecasted" predictions of the baseline GAMs provided an assessment of how 2021 data may have deviated from historical and normal seasonal variation.  The mgcv R package was used to create the models [@Wood17].

@@

Quantitative assessments of trends included descriptive summaries, Spearman rank correlations between pairs of variables, and multiple comparison tests to assess trends between months.  Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.  Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a. All variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.  Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.

For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners. Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and evaluated with methods described below. Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.

When appropriate, methods were used from the NADA R package [@Lee20] to accommodate measured concentrations in water quality variables that were below detection.  These included estimates of median, mean, and standard deviation of parameters using the "cenfit" function from NADA.  For the multiple comparison tests, only total nitrogen, secchi, and chl-a data were evaluated, none of which were below detection as noted in the FDEP data.  Nitrate/nitrite concentrations were not evaluated quantitatively because over 80\% of the observations were below detection.  Correlation analyses were based on weekly aggregations by area, using appropriate methods for median estimates with non-detects in the water quality data.

@@

Pairwise Spearman rank correlations showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).  Among the water quality variables, chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH ($p < 0.005$ for all) and negatively correlated with secchi depth ($p < 0.05$). Total nitrogen was positively correlated with total phosphorus, temperature, and ammonia ($p < 0.005$, except $p < 0.05$ for temperature) and negatively correlated with secchi ($p < 0.005$).  Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas red macroalgae was negatively correlated with cyanobacteria ($p < 0.005$).  Among the seagrasses, manatee grass was negatively correlated with shoal grass and turtle grass ($p < 0.005$).  Several notable correlations were also observed between water quality, macroalgae, and seagrass. Chlorophyll-a was negatively correlated with red and green macroalgae ($p < 0.05$), whereas secchi depth was positively correlated with both ($p < 0.05$) and negatively correlated with cyanobacteria ($p < 0.05$).   For water quality and seagrass, chl-a, total nitrogen, and total phosphorus were negatively correlated with turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively correlated with shoal grass ($p < 0.05$).  For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).

@@

Several of the results support known responses of water quality to nutrient loads for shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].  Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations (Figure \@ref(fig:allcor)). Temporally, these changes were observed at different times and for different species of phytoplankton.  The initial increase in chlorophyll-a was first associated with a diatom bloom in April and then by *K. brevis* in June and July.  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where total nitrogen that included organic sources were at high concentrations through April and again peaking in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].

From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability during the summer 2021 in comparison to prior years. Future effects on seagrass persistence in this region of Tampa Bay are unknown at this time, but monitoring continues.  Changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous and not readily apparent over the initial 4 month sampling period.  Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.  Spatially, transects closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay (area 3) had higher frequency occurrence of turtle grass and lower frequency of shoal grass.  This reflects a natural partitioning of the species along a salinity gradient [@Sherwood17].  Correlations between macroalgae and seagrasses can also be explained by spatial differences along a depth gradient. Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.  Shoal grass and turtle grass generally grow at more shallow depths than manatee grass and are often exposed during low tides.  As an adaptive mechanism to varying water levels at shallower depths, blades of shoal grass and turtle grass can "lay down" during low tide to minimize desiccation stress.  Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species grows deeper. In addition to species depth partitioning, these associations may also reflect natural variation in seagrass response to water clarity gradients along the main axis of Tampa Bay.  Shoal grass can occur in more turbid waters of oligohaline areas.

@@
